---
title: "Class 10: Halloween Candy Mini project"
author: "Charlize Molitor (PID: 18515740)"
format: pdf
toc: TRUE
---
## Background
As it is nearly Halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy!

Our data comes from the 538 website and is available as a CSV file: 

## Data Import

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

```{r}
flextable:: flextable(head(candy))
```



> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

```{r}
candy |>
  nrow()
```

```{r}
library(tidyverse)

candy %>%
  nrow()
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. My favorite candy and it's winpercent

```{r}
candy["Reese's Peanut Butter cup",]$winpercent
```

```{r}
library(dplyr)

candy |>
  filter(rownames(candy)== "Reese's Peanut Butter cup") |>
  select(winpercent)
```

 > Q4. What is the winpercent value for “Kit Kat”?
 
```{r}
library(dplyr)

candy |>
  filter(rownames(candy)== "Kit Kat") |>
  select(winpercent)
```
 
> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
library(dplyr)

candy |>
  filter(rownames(candy)== "Tootsie Roll Snack Bars") |>
  select(winpercent)
```


## Quick over view of the dataset

```{r}
library(skimr)

skimr::skim(candy)
```

 > Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
 
 The winpercent is on a 0-100 scale and the rest are 0-1 scale
 
 
> Q7. What do you think a zero and one represent for the candy$chocolate column?

That the candy does not contain any chocolate 


 > Q8. Plot a histogram of winpercent values
 
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 20)
```
 
> Q9. Is the distribution of winpercent values symmetrical?

No, not symmetrical based on the graph. 

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```


> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate is higher ranked than fruit candy

```{r}
# 1. Find all chocolate candy in the dataset
# 2. Find their winpercent values
# 3. Calculate the mean of these values

# 4-6. Do the same for the fruity candy
# 7. Compare mean winpercent of chocolate vs fruity
# 8. Pick the highest as the winner


choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds,]$winpercent
choc.mean <- mean(choc.win)
choc.mean
```

```{r}
fruit.inds <- candy$fruity==1
fruit.win <- candy[fruit.inds,]$winpercent
fruit.mean <- mean(fruit.win)
fruit.mean
```


> Q12. Is this difference statistically significant?

There is a staticical significance when to comes to chocolate and fruity candy

```{r}
t.test(choc.win,fruit.win)
```

## Overall Candy rating

 > Q13. What are the five least liked candy types in this set?
 
```{r}
library(dplyr)

candy |>
  arrange(winpercent) |>
  head(5)
```
 
 
```{r}
x <- c(5,10,15,4)
#sort(x)
order(x)
#arrange(candy$winpercent)
```
 
```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,], 5)
```
 
 >Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```

 > Q15. Make a first barplot of candy ranking based on winpercent values.
 
```{r}
library(ggplot2)


ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
 
 > Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
 
```{r}
ggplot(candy) +
  aes(x = winpercent, 
      y= reorder(rownames(candy), winpercent)) +
  geom_col()
```
 > Q17. What is the worst ranked chocolate candy?
 
 One dime
 
> Q18. What is the best ranked fruity candy?

Starburst

Add some color based on the type of candy

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "pink"
my_cols[as.logical(candy$fruity)] <- "blue"
my_cols[as.logical(candy$bar)] <- "purple"
my_cols
```

```{r}
ggplot(candy) +
  aes(x = winpercent, 
      y= reorder(rownames(candy), winpercent)) +
  geom_col(fill= my_cols)
```
 
 ## Winpercent and Pricepercent
 
 A plot with both variables/columns winpercent and pricepercent 
 
```{r}
my_cols[as.logical(candy$fruity)] <- "blue"

ggplot(candy) +
  aes(x= winpercent, 
      y= pricepercent, 
      label=rownames(candy)) +
  geom_point(col= my_cols) +
  geom_text()
```
 
```{r}
library(ggrepel)

ggplot(candy) +
  aes(x= winpercent, 
      y= pricepercent, 
      label=rownames(candy)) +
  geom_point(col= my_cols) +
  geom_text_repel(max.overlaps = 7)
```
 
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )

```




> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?


```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
 
 
 
 
 
 
 ## Exploring the correlation structure
 
 Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.
 
```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```
 
  > Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
  
Chocolate and fruit candies, mixed together. 
  
>Q23. Similarly, what two variables are most positively correlated?
 
 Chocolate with chocolate candies. 
 
 
## Principle Component Analysis

The function to use it called `prcomp()` with the optional `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Our main PCA results figure

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols) 
```

We should also examine the variable "loading's" or contributions of the original variables to the new PCs

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols) 
```

Interative plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interative and will not render to PDF :(
```{r}
library(plotly)

##plotly(p)
```

