---
title: "Class 17"
author: "Charlize Molitor (PID:18515740)"
format: pdf
---

## Section 1: Proportion of G|G in a population 

Downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39835117-39955118;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```


Now lets look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population 

Let's now dig into this further. 

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. 

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```


```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```


Let's make a boxplot

```{r}
ggplot(expr) +
  aes(geno,exp, fill = geno) +
  geom_boxplot(notch = TRUE)
```

 > Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.

Based on the boxplot, there is less of the G|G gene than the A|A. The SNP would effect the ORMDL3 




